Application of Micro-Rotors to Induce Crystal Nucleation: Challenges, Takeaways, and Future Prospects\\ PhD Thesis

### 2025-04-15 Latex_files/Thesis/chapters/chapter 5/main.tex
updates 15/04/25

\chapter{Detection and Characterisation of rotational spherical aggregate rotational dynamics}
\label{chapter:simulated_detection}
As outlined in the end of Chapter 4, one of the difficulties 
in characterising interactions with asymmetric objects is 
the coupled motion between translation and rotation. In order
to characterise the optical trap a position detection system 
such as a lateral effect photodiode or a quadrant photo diode
(see fig.\ref{fig:position_detectors}). Typically, a position 
detection system assumes that there exists a linear relationship 
between the a particle's displacement and the detected signal.
However, this is not always the case as demonstrated in chapter
4; as dimer's demonstrate non-harmonic trapping beyond a certain 
displacement. In addition, the rotational effects of the trapped
dimer become a more significant factor when characterising the 
trap strength. This can have unintended effects when it comes to
areas of research that require a precise understanding of the 
hydrodynamic behaviour of particles. As such there has been a 
concerted effort to either eliminate rotational motion, or describe
how the rotational behaviour influences the trajectory reported 
by the position detector.

In chapter 4 we showed that while machine learning can be used 
to accurately predict the translational displacements of an 
aggregate particle, it fails when trying to account for the 
rotational behaviour of said particle. In this chapter we 
build upon \cite{BarZiv1997} but now account for the full 
rotational motion. This was initially developed in tandem with 
the work conducted in chapter 3, with the idea being that if a 
crystal nucleus formed on a rotating sphere how would that 
nucleus influence the motion of the rotating sphere. Using 
static light scattering we can map the outputted signal from a 
optically trapped dimer to its expected orientation in real 
time. Thus allowing us to estimate the optical torque being 
applied to the dimer by the trapping beam. As demonstrated 
in chapter 4, the influence of a circularly polarised beam 
can result in gyroscopic precession for specifically sized 
dimers.

We start by describing a quadrant photo diode system 
simulated the power spectrum derived from a particle 
trajectory. We demonstrate that even in idealised 
circumstances the rotational motion is not revealed 
from the collected power spectrum. We then go about 
defining an optical fibre detection system from a 
mathematical perspective. Before demonstrating how it 
can be used to instantaneously predict orientational 
information on a trapped particle. We then consider 
common factors of error in the characterisation process, 
such as signal error, and incorrect particle sizing. 
Both of which have a significant impact on the performance 
of the model, but by utilising Bayesian inference, and 
time averaging we can better refine our model.

\section{Characterisation of asymmetric dimer dynamics via PSD analysis}
As discussed in \ref{sec:simulated_QPD}, one of the methods 
developed to work in conjunction with \cite{Vigilante2020} 
is a simulated quadrant photo diode for as a position 
detection system. While it is possible to extract all of 
the relevant dynamical information from a simulation, 
confirming the same behaviour in an experimental setting 
can be challenging if dealing with a non-birefringent 
anisotropic scatterer. The QPD is composed of 4 photodiodes 
that measure the intensity of light incident on their 
surfaces. Using \textit{ott} we can define a region in the 
far-field that is analogues to the surface of the QPD surface. 
Evaluating the electric field across this surface gives us 
an approximation of the QPD signal outputted in an experimental 
situation. This does not provide a one-to-one result however 
as hardware errors (i.e. internal resistance, external light 
sources, and vibrations) would distort the signal somewhat. We 
can use this as a means to examine the limitations of using 
back-focal plane interferometry as a means for characterising 
anisotropic scatterers.

As a benchmark we start by considering a single sphere within 
an optical trap. A single polystyrene sphere suspended in water 
($a=1\mu m$, $n_p=1.59$, $n_m=1.33$) was trapped by a focused 
Gaussian beam ($NA=1.25$) using circularly polarised light. For 
the sake of time efficiency the trajectory was sampled every 10
time steps, meaning the upper bound on the power spectra is
$f_{Nyq}=f_{sample}/2=5000\ Hz$. To optimise the frequency window
we fitted the power spectra using the aliased Lorentzian 
(Eq.~\eqref{eq:aliased_lorentzian}). 
\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth]{PSD_sphere.png}
	\caption{Recorded power spectra fitted to eq.~
		\ref{eq:alaised_lorentzian}, scattered points represents 
		the blocked data ($n_b=100$). Corner frequency for the 
		Lorentzian curves are reported in the legend.}
	\label{fig:psd_sphere}
\end{figure}

@@

The reason for the discrepancies between all 3 methods is due 
to what is actually being measured. The \textit{ott} estimates 
are simply looking at the differences in the trapping strength 
along the Cartesian axis'. If calibrated over an long enough 
time frame you would expect that the resulting power spectra 
would exactly mirror the \textit{ott} predictions. However over 
a short calibration time a QPD will only sample a small trajectory 
that does not fully explore the trapping plane. There is a clear 
trade off in terms of accuracy and computation time as shorter 
calibration runs are computationally more efficient but prone to 
errors.

With this in mind, let us consider a symmetric dimer that is 
optically trapped by the same Gaussian beam. Not only does 
the dimer's equilibrium position change but it is subjected 
to rotational motion due to its unequal moments of inertia. 
This is reflected in the calibration results using the simulated 
QPD, where we see a drastically different estimation between the 
\textit{ott} estimate and the QPD estimate.

\begin{center}
	\captionof{table}{QPD fitting for symmetric dimer}
	\label{tab:dimer}
	\begin{tabular}{ |c|c|c|c|c|c|c| } 
		\hline
		Fitting parameter & \multicolumn{2}{|c|}{\textit{ott} estimate} & 
		\multicolumn{2}{|c|}{QPD fitting} & \multicolumn{2}{|c|}{Trajectory fitting} \\
		\hline
		$f_c\ [Hz]$ & 445 & 409 & \parbox{1cm}{\centering 431} & 424 
		& \parbox{1.25cm}{\centering 274} & 285 \\
		$\sigma(f_c)\ [Hz]$ & --- & --- & 9.22 & 9.16 & 7.82 & 7.91  \\
		$\kappa\ [pN/\mu m]$ & 52.82 & 48.54 & 51.13 & 50.26 & 32.45 & 33.75 \\
		\hline
		Ellipticity &
		\multicolumn{2}{|c|}{28.5 \%} &
		\multicolumn{2}{|c|}{12.7 \%} & 
		\multicolumn{2}{|c|}{13.8 \%} \\
		\hline
	\end{tabular}
\end{center}

@@

\subsection{QPD for angular displacement detection}
The next logical step is to consider whether or not a QPD can at 
all be utilised for detection of rotational motion. This has been
shown for nano-particles that exhibit periodic rotational motion
using by considering the difference in diagonal quadrants. The 
benefit of detecting rotational motion is that one can begin to 
build a better understanding about how angular momentum is 
transferred to a dimer which could be extended to more complex
particles.

At just a cursory glance it would appear that there is a clear
relationship between the QPD signal and the orientation of the 
dimer. If we look at the QPD signal produced by a dimer as it is 
rotated we can see that it is equally sensitive to rotations in
either the X-Z or Y-Z plane but there is no change in the signal 
when rotated by in the X-Y plane.

@@

The dependence of $S_x$ and $S_y$ and the dimer's orientation
(defined by the spherical angles $\theta$, $\phi$) is therefore
not clear. To that end, we utilised machine vector regression,
which takes as an input the voltage from the 4 quadrants and 
tries to fit that to the dimer's trajectory. Fitting the QPD
signal to the dimer's position shows promising results, shown in
fig.~X.X is a prediction of a symmetric dimer's position based
on the QPD signal. With the actual displacement on the x-axis 
and the predicted result y-axis, and the dashed line represents 
the y=x line - indicating an ideal prediction.

Fig~X.X shows that its relatively trivial to predict a particle's
position based off the QPD signal, even in axial direction which
has not been done previously. It should be noted that these 
displacements are relative to the trap focus and so accurate 
tracking of the beam movements needs to be taken. This result is
partially due to the fact that displacement and signal units are
linearly related by \eqref{eq:conversion_factor}. Interestingly 
when we run the same protocol for different sized dimer's there 
seems to be a cut off point when the fit begins to fail. Plotting
the $R^2$ for each fit vs its size ratio shows that beyond a size
ratio of 5 the machine learning begins to fall off significantly.

This is surprising because we would expect that as the second 
sphere shrinks the dynamics should more closely approximate that 
of a single sphere. We can compare our

@@

\subsection{Dimer}
The dimer is defined by two spheres with refractive indices
of 1.59 and suspended in water ($n_{med} = 1.33$). \textit{MSTM}
requires a target origin for setting the scattering expansion, 
there is no agreed upon solution for where the target origin 
should be relative to the sphere positions. In chapter 4 we showed
that altering the target origin would drastically alter the 
tmatrix produced

@@

At the same we use \textit{mstm} to replicate the intensity of light
incident on the surfaces of detectors placed around the focus of the
optical trap. The exact number of detectors was initially assumed to 
be arbitrary, in that it made no difference to our estimate whether 
we used 2 angles or 200. For practical purposes it seemed beneficial 
that we demonstrate our method works for a minimal number of detection 
angles, as geometric constraints come into play when trying to install 
a high number of detection fibres for any optical tweezer set up.

>>>>

\chapter{Detection of rotational spherical aggregate rotational dynamics}
\label{chapter:simulated_detection}
As outlined in the end of Chapter 4, one of the difficulties 
in characterising interactions with asymmetric objects is 
the coupled motion between translation and rotation. In order
to characterise the optical trap a position detection system 
such as a lateral effect photodiode or a quadrant photo diode
(see fig.\ref{fig:position_detectors}). Typically, a position 
detection system assumes that there exists a linear relationship 
between the a particle's displacement and the detected signal.
However, this is not always the case as demonstrated in chapter
4; as dimer's demonstrate non-harmonic trapping beyond a certain 
displacement. In addition, the rotational effects of the trapped
dimer become a more significant factor when characterising the 
trap strength. This can have unintended effects when it comes to
areas of research that require a precise understanding of the 
hydrodynamic behaviour of particles. As such there has been a 
concerted effort to either eliminate rotational motion, or describe
how the rotational behaviour influences the trajectory reported 
by the position detector.

In this chapter we build upon the work of \cite{Rohrbach2002} and
\cite{BarZiv1997}; the former used to measure the relative position
of the trapped aggregate, and the latter to measure the orientation
of the aggregate. This was initially developed in tandem with the 
work conducted in chapter 3, with the idea being that if a crystal 
nucleus formed on a rotating sphere how would that nucleus influence 
the motion of the rotating sphere. Using static light scattering we 
can map the outputted signal from a optically trapped dimer to its 
expected orientation in real time. Thus allowing us to estimate the 
optical torque being applied to the dimer by the trapping beam. As demonstrated in chapter 4, the influence of a circularly polarised 
beam can result in gyroscopic precession for specifically sized 
dimers.

We start by describing a quadrant photo diode system simulating the 
power spectrum derived from a particle trajectory. We demonstrate 
that even in idealised circumstances the rotational motion is not 
revealed from the collected power spectrum. We then go about 
defining an optical fibre detection system from a mathematical 
perspective. Before demonstrating how it can be used to 
instantaneously predict orientational information on a trapped 
particle. We then consider common factors of error in the 
characterisation process, such as signal error, and incorrect particle 
sizing. Both of which have a significant impact on the performance 
of the model, but by utilising Bayesian inference, and time averaging 
we can better refine our model.

\section{Characterisation of asymmetric dimer dynamics via PSD analysis}
As discussed in \ref{sec:simulated_QPD}, one of the methods 
developed to work in conjunction with \cite{Vigilante2020} 
is a simulated quadrant photo diode for as a position 
detection system. While it is possible to extract all of 
the relevant dynamics from a simulation, from an experimental
standpoint characterising those dynamics is dependent on what
experimental techniques are used. Translational motion can be
characterised via a position detection system but angular 
motion is far more difficult to detect, let along characterise.

The simulative QPD is composed of 4 photodiodes that measure the 
intensity of light incident on their surfaces. Given that the 
condensing lens will have a maximum acceptance angle given by 
its numerical aperture ($\theta_{max} = \sin^{-1}\left(NA_c/n_
{med}\right)$) we define the electric field incident on the QPD 
surface as the elements of the scattered and incident field that 
propagate within that angular range of $\pm \theta_{max}$. Using 
\textit{ott} we can evaluate the total electric field to an 
arbitrary spacial resolution across the surface of the QPD. The 
reported signal is found by taking the difference between pairs 
of quadrant signals (see \ref{sec:simulated_QPD} for further 
detail). This does not provide a one-to-one result however as 
hardware errors (i.e. internal resistance, external light sources, 
and vibrations) would distort the signal. It can be used however to 
show how if the reported power spectra is an accurate representation 
of the true particle dynamics or if the additional degrees of freedom 
have an adverse effect on the scattering. See chapter 2 for a breakdown
of power spectral analysis.

As a benchmark we start by considering a single sphere within an optical 
trap. A single polystyrene sphere suspended in water ($a=1\mu m$, 
$n_p=1.59$, $n_m=1.33$) was trapped by a focused Gaussian beam ($NA=1.25$) using circularly polarised light. For the sake of time efficiency the trajectory was sampled every 10 time steps, meaning the upper bound on 
the power spectra is $f_{Nyq}=f_{sample}/2=5000\ Hz$. To optimise the 
frequency window we fitted the power spectra using the aliased Lorentzian.
(Eq.~\eqref{eq:aliased_lorentzian}). 
\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth]{PSD_sphere.png}
	\caption{Recorded power spectra fitted to eq.~
		\ref{eq:alaised_lorentzian}, scattered points represents 
		the blocked data ($n_b=100$). Corner frequency for the 
		Lorentzian curves are reported in the legend.}
	\label{fig:psd_sphere}
\end{figure}

@@

The reason for the discrepancies between all 3 methods is due 
to what is actually being measured. The \textit{ott} estimates 
are simply looking at the change in the trapping force as you
move along a given axis'. In reality, a particle will be freely
diffusing within the trap focus meaning over a short calibration 
time the QPD is only estimating a weighted average of the trapping
strength. If calibrated over an long enough time frame you would 
expect that the resulting power spectra would exactly mirror the 
\textit{ott} predictions. There is a clear trade off in terms of 
accuracy and computation time as shorter calibration runs are 
computationally more efficient but prone to errors. However, it
should be noted that the estimation made by the QPD is still less
than the results from \cite{BergSoerensen2004}, though this could
be due to a lack of external noise signals.

With this in mind, let us consider a symmetric dimer that is 
optically trapped by the same Gaussian beam. Not only does 
the dimer's equilibrium position change but it is subjected 
to rotational motion due to its unequal moments of inertia. 
This is reflected in the calibration results using the simulated 
QPD, where we see a drastically different estimation between the 
\textit{ott} estimate and the QPD estimate.
\begin{center}
	\captionof{table}{QPD fitting for symmetric dimer}
	\label{tab:dimer}
	\begin{tabular}{ |c|c|c|c|c|c|c| } 
		\hline
		Fitting parameter & \multicolumn{2}{|c|}{\textit{ott} estimate} & 
		\multicolumn{2}{|c|}{QPD fitting} & \multicolumn{2}{|c|}{Trajectory fitting} \\
		\hline
		$f_c\ [Hz]$ & 445 & 409 & \parbox{1cm}{\centering 431} & 424 
		& \parbox{1.25cm}{\centering 274} & 285 \\
		$\sigma(f_c)\ [Hz]$ & --- & --- & 9.22 & 9.16 & 7.82 & 7.91  \\
		$\kappa\ [pN/\mu m]$ & 52.82 & 48.54 & 51.13 & 50.26 & 32.45 & 33.75 \\
		\hline
		Ellipticity &
		\multicolumn{2}{|c|}{28.5 \%} &
		\multicolumn{2}{|c|}{12.7 \%} & 
		\multicolumn{2}{|c|}{13.8 \%} \\
		\hline
	\end{tabular}
\end{center}

@@

\subsection{QPD for angular displacement detection}
The next logical step is to consider whether or not a QPD can at 
all be utilised for detection of rotational motion. This has been
shown for nano-particles that exhibit periodic rotational motion
using by considering the difference in diagonal quadrants 
\cite{Yifat2021}. The only difference between these results is 
the fact that rotation occurs perpendicular to the particle's long 
axis, whereas dimers rotate about their long axis. The benefit of 
detecting rotational motion is that one can begin to build a better understanding about how angular momentum is transferred to a dimer 
which could be extended to more complex particles. This could give 
some understanding to the results of Chapter 4.

At just a cursory glance it would appear that there is a clear
relationship between the QPD signal and the orientation of the 
dimer. If we look at the QPD signal produced by a dimer as it is 
rotated we can see that it is equally sensitive to rotations in
either the X-Z or Y-Z plane but there is no change in the signal 
when rotated by in the X-Y plane. 
\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{rotation_figure}
	\caption{QPD signals detected by a dimer being rotated while
			centred at the focus of the optical trap. Top right:
			rotation in the X-Z plane. Top left: rotation in the
			Y-Z plane. Bottom: Rotation in the X-Y plane.}
	\label{fig:QPD_rotation}
\end{figure}

@@

The dependence of $S_x$ and $S_y$ and the dimer's orientation
(defined by the spherical angles $\theta$, $\phi$) is therefore
not clear. To that end, we utilised machine vector regression,
which takes as an input the voltage from the 4 quadrants and 
tries to fit that to the dimer's trajectory. Fitting the QPD
signal to the dimer's position shows promising results, shown in
fig~\ref{fig:QPD_rotation} is a prediction of a symmetric dimer's 
position based on the QPD signal. With the actual displacement 
on the x-axis and the predicted result y-axis, and the dashed 
line represents the y=x line - indicating an ideal prediction.

Fig~\ref{fig:QPD_rotation} shows that its relatively trivial 
to predict a particle's position based off the QPD signal, even 
in axial direction which has not been done previously. It should 
be noted that these displacements are relative to the trap focus 
and so accurate tracking of the beam movements needs to be taken. 
This result is partially due to the fact that displacement and 
signal units are linearly related by \eqref{eq:conversion_factor}. Interestingly when we run the same protocol for different sized 
dimer's there seems to be a cut off point when the fit begins to 
fail. Plotting the $R^2$ for each fit vs its size ratio shows 
that beyond a size ratio of 5 the machine learning begins to fall 
off significantly.

This is surprising because we would expect that as the second 
sphere shrinks the dynamics should more closely approximate that 
of a single sphere. This should be reflected in the total field
incident on the QPD and therefore the machine learning program 
should be able to easily detect the dimer's translational motion.
We can therefore only conclude that the dimer's rotational motion
is contributing to the scattering to the scattered field, even if
it is miniscule the contribution is enough to throw off a 
position detection system. Detection of rotational motion was 
attempted using machine learning, the model was trained on the 
same trajectory mentioned in \ref{sec:off-axis}, where the dimer 
is trapped in an off-axis orientation, from a top down perspective
it should be easy enough to note that the dimer is not vertically 
trapped even if the exact angle is difficult to determine.

Overall, the reliance on machine learning to perform vector 
regression on the collected scattered field is not a viable 
method for detecting let along characterising the angular 
displacement of an isotropic scatterer. While there is a 
clear correlation in the scattered signal and angular 
displacement the translational motion makes up the majority 
of the expected signal detected by the QPD. As of now, there
is no optical arrangement that would allow rotational motion 
while restricting translational motion.

@@

\subsection{Dimer}
The dimer is defined by two spheres with refractive indices
of 1.59 and suspended in water ($n_{med} = 1.33$). \textit{MSTM}
requires a target origin for setting the scattering expansion, 
there is no agreed upon solution for where the target origin 
should be relative to the sphere positions. In chapter 4 we showed
that altering the target origin would drastically alter the 
tmatrix produced. We choose (somewhat arbitrary) to use 
the dimer's centre of diffusion which is located $\zeta a_I$ from 
the centre of the larger sphere. For a discussion see \emph{X.X}.

@@

At the same we use \textit{mstm} to replicate the intensity of light
incident on the surfaces of detectors placed around the focus of the
optical trap. The exact number of detectors was initially assumed to 
be arbitrary, in that it made no difference to our estimate whether 
we used 2 angles or 200.
